---
title: "Overview of Cost Estimation Methods"
author: "James Woods"
date: ""
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Prelude

+ Costs are always measured with some error.
+ Not all cost forecasts are good.
+ The incentives to reduce costs vary widely.

## How do you mismeasure costs?

+ Misclassify regulated vs non-regulated
    + Incentives to do this.
+ Cost categories (Account codes)
    + Insurance moves from selling expenses to administrative
+ Reallocation from one customer type to another.
    + Industrial to residential, for example if a new subdivision or two springs up.
+ To and from Rate Base
    + Costs flow through but Rate base gets a return.
    
## Random events.
  
  + Cost increasing random events are more frequent than cost saving.
  + Extraordinary items are on income statement.
  + Random events are more likely to increase costs than lower them.

## General Methods

+ Base or Test year
+ Ideal system
+ Regression
+ Corrected Regression
+ Frontier Regression
+ Data Envelopment
+ Yardstick

In almost all cases, the idea is to find out what the revenue requirement should be and after that setting prices, tariffs, to meet that requirement given other objectives.

## Test Year

+  Use the costs and sales volume from this year to develop revenue requirements and tariff structure for the next few years.
    + One data point is used (under one interpretation)

+ *Adversarial system*: Utilities on one side and rate payer advocate (Often commissioners staff) on the other and commissioners acting as judge.
    + Fight over cost, rate base, share holder allocation of costs.
    
+ *Accounting system*: Auditable with receipts.  It is what really happened.

## Ideal System

+ Observe the needs of the system, the location of the customers and what they consume.
+ Design a system that would satisfy these customers.
+ Price that system.
+ Example: Oregon Quality Education Model (http://www.ode.state.or.us/search/results/?id=166)
+ You can see other examples of this in teh EIA cost estimates.

## Ideal System Comments

+ Very hard to estimate costs.  Common numbers
    + Rough: within 30-60% and biased low
    + Semi-detailed: within 20-30% and biased low
    + Detailed (with blueprints and bids): 3-5% and biased low

+ They run low because all cost estimates run low.  You don't accidentally add in something expensive but you do forget to add in something expensive.
+ Ideal systems don't have legacy systems or history.
    + Example of SMUD and Rancho Seco in 1966, which closed in 1989.
+ All these cost estimates are very expensive.    
    
    
## Multi Year (Period) Methods
  
  + The next group:
      + Regression
      + Corrected Regression
      + Frontier Regression
      + Data Envelopment Analysis (DEA)
  + Use comparable data from many time periods.
  + I will demonstrate with artificial data.

Please note that even in the test year methodology you still use these techniques behind the scenes.
  
## Fake Data

+ I will create data of the form $Cost = F + \beta kWh_t + e_t +\nu_t$
    + $F$ is intended to be a fixed cost (\$ 5000)
    + $\beta$ is a constant marginal cost (\$ 0.05 )
    + $kWh_t$ is the kWh generated in time t.
    + $e_t$ is a mean zero uncertainty in time t.
    + $\nu_t$ is uncertainty that is positive and bias up.  Accidents.
    
+ More could be done with increasing cost of generation and different costs of producing at different times of day and by different generation units.

## Data Summary
```{r DataGen, include=FALSE}

FakeData <- data.frame(kWh = runif(100, min = 1000, max = 8000))
FakeData$e <- rnorm(100, mean = 0, sd = 100)
FakeData$nu <- rexp(100, rate = .01)
FakeData$Cost <- 5000 + .05 * FakeData$kWh + FakeData$e + FakeData$nu

```

```{r Table}
library(knitr)
kable(summary(FakeData))

```

  
## The Data  

```{r DataPlot, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
FakeData %>% ggplot(aes(kWh, Cost)) + geom_point() -> BaseGraphics

BaseGraphics
```

## Regression

+ The idea of a regression is to run a function through the dots such that the sum of squares of the vertical distances is as small as possible.

-------
```{r Regression,  echo=FALSE, message=FALSE, warning=FALSE , results = "asis"}
library(stargazer)

RegressionModel <- lm(Cost ~ kWh, data=FakeData)
RegressionModel %>% stargazer(header = FALSE)

```


##
```{r}
BaseGraphics + geom_smooth(method = "lm")

```


## Comments on Regression

+ Will generally hit the true slope parameters (If the volume is measured without noise) but not the intercept.
    + When it is correctly specified, right parameters and shape.
+ Yes, there is uncertainty.
+ One observation can't do much but you can measure the influence and see the effect easily.

## Corrected Regression

+ Slides the regression line so that all the dots are above the line.
+ Intercept term is $intercept + min(e +\nu)$
+ Intended to be the ideal, what can be accomplished.

##

```{r}
BaseGraphics + geom_abline(intercept = coef(RegressionModel)[1] + min(resid(RegressionModel)), slope = coef(RegressionModel)[2])

```

## Corrected Regression
  
+ One observation moves the line.
+ Always understates fixed cost.
+ Makes statisticians sigh.

## Frontier Regression

+ Regression assumes that the errors, deviations from the line, are symmetric, same amount above as below.
+ Frontier says the errors can by asymmetric
+ Commonly assume a non-normal distribution like the gamma. We used normal-exponential (Special kind of gamma) mixture

##

```{r Gamma,  echo=FALSE, message=FALSE, warning=FALSE , results = "asis"}
library(stargazer)
library(frontier)
sfa(Cost ~ kWh, data=FakeData, ineffDecrease = FALSE) -> Frontier

```

```{r, echo = FALSE, results='markup'}

Frontier %>% summary()
```

##
```{r}
BaseGraphics + geom_abline(intercept = coef(Frontier)[1] , slope = coef(Frontier)[2])+ geom_smooth(method = "lm")

```

## Comments on Frontier

+ This is how we generated the data 
+ Not as sensitive to one observation point
+ Gives an estimate of the potential efficiency gains.

## Data Envelopment

+ Create the smallest shape around the bottom of the dots drawing from dot to dot.
+ Convex hull is the technical term.
+ Efficiency is measured as distance from the frontier
+ This one looks stupid.  The axes are reversed because of technical limitations.  Shows the highest kWh per cost level.

##
```{r DEA, message=FALSE, warning=FALSE}
library(Benchmarking)
dea.plot.frontier(FakeData$Cost,FakeData$kWh, xlab="Cost", ylab = "kWh" )
```

## Comments on DEA

+ Preferred by engineers
+ Some variations, bootstrap, give uncertainty.
+ See how it depends on only a few observations.
+ You can also use it for isoquants.

## Yardstick

+ This means look at the costs of other firms.
+ The focus is not on accuracy but on incentives to tell the truth and to reduce costs.
    + If every cost reduction you make results in lower prices and no benefit to shareholders, why do it?
    + With yardstick, if you make a cost reduction before the other guy does, your shareholders gain.
+ Can be combined with the other techniques.    
    