---
title: "Cost Measurements (Ch 5)"
author: "James Woods"
date: ""
output:
  beamer_presentation: default
  ioslides_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```



# Depreciation


## Review from Intro to Costs

Assets cost a lot and while purchased in one year provide benefits over many years. We would like to allocate cost of purchasing that asset over the years we use it (Matching principle).


+ When you buy the asset (Balance Sheet) it is in the rate base and earns interest
+ Every year some of the purchase price turns into depreciation (Income statement) and passed on the consumers.
+ Ideally, sale of an asset should result in no income changes but it almost always happens.

## Extended Numerical Example

+ Data Needed to Depreciate an Asset
    + Service Date
    + Cost Basis
    + Asset Life
    + Salvage Value
    + Method, accelerated, straight line, etc.

## Example

+ Asset goes into service now.
+ Cost 10K
+ Will last five years
+ Salvage for 2K
+ Tax uses MACRS but we will use straight line.

Depreciation each year:

$$\frac{(Cost - Salvage)}{Life} = \frac{10K - 2K}{5} = `r 8/5`K$$
The book spends a few pages doing this in percentage terms but that is more about the "Iowa Curve" methodology, what survival analysis looked like in the 1950s

## Example Continued

```{r, include=FALSE}
Dep <- data.frame(Year=c(0,1,2,3,4,5), Depreciation = 8/5, Book = 10)
Dep$Depreciation[1] <- 0
for(i in 2:6) Dep$Book[i] <- 10 - sum(Dep$Depreciation[1:i]) 

```

```{r}
library(knitr)
kable(Dep)
```

The Book Value is the cost basis less accumulated depreciation.  This is in the rate base and earns the allowable return.

## Example to Revenue Requirement

```{r, include = FALSE}
Dep$RR <- Dep$Depreciation + .1*Dep$Book
Dep$RR[1] <- 0 

```

All the depreciation is part of the revenue requirement and with an allowable rate of return of 10\%, 10\% of the book value.

```{r}
library(knitr)
kable(Dep)
```

## Asset Retirement Obligations

Now lets add in that a site will need to be cleaned up or that it is expensive to decommission.  In this case 5K.

  + Every year, take an equal amount as an expense.
  + Add to the rate base, it earns interest in the rate base.
  + When you decommission, remove it from the rate base but don't add it as an expense since you already did.

```{r, include =FALSE}
Dep$RR <-NA
Dep$Decomission <- 1
Dep$Decomission[1] <- 0

Dep$ARO <-0
for(i in 2:6) Dep$ARO[i] <- sum(Dep$Decomission[1:i]) 


Dep$RR <- Dep$Depreciation + Dep$Decomission + .1*(Dep$Book+Dep$ARO)
Dep$RR[1] <- 0 

```

```{r}
library(knitr)
kable(Dep)
```

# Disallowance

## Examples of Dissallowance and the Reasons Why

Disallowance is the technical term for not including an expense or an asset in the rate base.

Key terms:

+ Used and Useful: Comes from Smyth v. Ames (1898) and says physically used and useful to current rate payers before they can be asked to pay for it.  There are some exceptions.
+ Prudent:  Makes sense at the time it was made.  Not penalized for not knowing the future.


## Examples of Dissallowance in Orgon

+ Trojan Nuclear Power Plant (PGE) http://www.puc.state.or.us/admin_hearings/key_puc_cases/95_322.pdf

+ Multiple Wind Projects (PacifiCorp) https://apps.puc.state.or.us/orders/2008ords/08-548.pdf


## Trojan


## PacifiCorp Wind Projects



# Regulated Rate of Return

## What we will look at

+ The effects of capital structure on the rate of return
+ The rarity of using actual structure and a framework for choosing one in a regulatory framework.
+ Technical problems:
    + Subsidiarys
    + Used assets
+ Three ways to estimate ROE and a critique.

## Earnings Volitility and Debt



## Why Assume a Structure Rather than Measure


## Criteria for an Assumed Structure

http://www.sciencedirect.com/science/article/pii/S1040619011001722


## Technical Issues

+ Subsidiaries
+ Used Assets

## Subsidiary
<!-- Parant company and subsidiary issues. -->
<!--   Who's data to use. -->
<!--   How transfered -->

## Used Assets

<!-- Buying used assets.,  How different in regulatory world and why -->
<!-- Purchase accounting adjustments -->


# Return on Equity (ROE)

## Usual Methods and an Unavoidable Assumption

Methods:
+ Discounted Cash Flow (DCF)
+ Capital Asset Pricing Model (CAPM)
+ Risk Premium Model (RPM)

Sitting under all of them is one of the efficient market assumptions. 


## Three Basic Kinds of Efficient Market Assumptions

Usefulness of information to predict future prices

+ Weak: All past price data will not help.  
+ Semi-Strong: Public data does not help
+ Strong: No data public or private can help

## Grampa Ish and Betting on Horses

+ You don't know who will win.
+ Calculate your own odds of each horse.
+ Bet the horse that is paying off more than you think it should.
+ You win more often if you have your grandson talk to the jockeys.

## Behavioral Finance 

+ Behavioral Economics has a brother.
+ There are plenty of anomalies
    + Monday Effect, French (1980)
    + January Effect, Roll(1983)
    + Firm Size, Banz(1981)
    + Mean Reversion, DeBondt and Thaler (1985)
    + Momentum, Jegadeesh (1990)
    
+ Anomalies get smaller after a paper about it is published.
+ Many critiques have to do with statistical details.

## Comparable Data

The intent is to have observations of a firm just like the one being regulated but without the effect of regulation.

+ The intent is a quasi-experimental control.
+ Comparability is easier to work with for populations but harder for an individual.
    + There is no perfect match
    + The Parable of the Cookie and Parent's Problem.

## DCM

<!-- DCM often uses a blend of historical and forecast with GDP growth. -->

## CAPM 
<!-- CAPM as regression with alpha and beta -->


## RPM